A significance test in forward stepwise regression

### 2014-03-21 paper/paper.tex
Paper: finished p-value calculation! Code: fwd_step uses Pg, group_lasso light edits, rerun some simulations.

\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}

@@

For various sparsity levels $k$, Algorithm~\ref{algo:fs} was run on
various data sets with \textit{steps} equal to $k$ and the ``average
power'' was calculated as the number of true variables in the active
set after $k$ steps divided by $k$. Nonzero coefficients had magnitude
in a range of multiples of $\gamma := \sqrt{2\log(G)}$, e.g. in $[1.4
\gamma, 1.6 \gamma]$. Results are shown in
Figure~\ref{fig:fwdstepsim}. After computing the
coherence of these design matrices, some calculations show the required
sparsity level to guarantee exact recovery in these situations is
about 2 or smaller, and the required nonzero coefficient magnitude is
likely in the range of 10 to 100 times $\gamma$. These simulations are far
from the stringent conditions required by theory to guarantee perfect
recovery, but the performance, while not perfect, may still be good
enough for some purposes.

@@

We now describe how to calculate our p-value, following the discussion of
group lasso in \cite{tests:adaptive}. For the rest of this section let
$g = \gstar$ be the index of
the group attaining the maximum on line 3 of Algorithm~\ref{algo:fs}.
%\lambda = \norm{X_g^Ty}_2/w_g$ be the value of the maximum, and
%$r = \text{rank}(X_g)$ be the rank of the maximizing group.
For a vector $u \in \real^p$, let $u_h$ denote the coordinates
of $u$ corresponding to the columns of group $h$ in the design matrix $X$.
We can rearrange the columns of $X$ to group these adjacently, so that
\[
u^TX^T = \begin{pmatrix} u_1^T X_1^T & u_2^T X_2^T & \cdots & u_G^TX_G^T \end{pmatrix}
\]
One step of the calculation will be to find an orthonormal basis for the
linear space $\vecsp_g = \{ u \in \real^p : u_g^T X_g^T y = 0, u_h = 0
\text{ for all } h \neq g \}$ so that we can project orthogonally to
this space. If $X_g$ is a single column and $X_g^Ty \neq 0$ (which
should be the case since $g$ maximizes the absolute value of this
quantity), then the
space is trivial and the desired orthogonal projection is the identity.
Otherwise, if $X_g$ has $p_g > 1$ columns the space $\vecsp_g$ will
generally have dimension $p_g-1$.
% have dimension $r-1$ for $r = \text{rank}(X_g)$. This holds except in
% degenerate cases such as $X_g = 0$ or $r \ll p_g$, but these cases are not
% likely to occur since $\norm{X_g^T y}_2/w_g$ is supposed to be maximized.
% We proceed by forming the $n \times p_g$ matrix
% $X_g - X_g X_g^T y y^T X_g/\norm{X_g^Ty}_2^2$ and performing Gram-Schmidt
% on the first $r-1$ columns. For precise theorems showing that this will
% generally work, see \cite{outerp:indep}.
We compute an orthonormal basis by Gram-Schmidt and form a
$p \times p_g$ matrix, which we denote $V_g$, by appending 0's in the
coordinates corresponding to all groups $h \neq g$ and to one column
in group $g$ (since Gram-Schmidt only produces $p_g-1$ columns).
With this we are ready to define the projection

\begin{equation}
  \begin{aligned}
    \label{eq:proj}
    P_g &= \Sigma XV_g (V_g^T X^T \Sigma X V_g)^\dagger V_g^TX^T \\
   \end{aligned}
\end{equation}
and the conditional variance

\begin{equation}
  \begin{aligned}
    \label{eq:cvar}
    \sigma^2_g &= y^TX_gX_g^T(I-P_g)\Sigma X_gX_g^Ty / (w_g^2 \norm{X_g^Ty}_2^2) \\
  \end{aligned}
\end{equation}
These simplify when $X_g$ is a single column, in which case $P_g = 0$
and $\sigma^2_g = X_g^T\Sigma X_g^T$.

Define $U_g = X_g^Ty/(w_g\norm{X_g^Ty}_2)$.
\begin{algorithm}
  \caption{Computing p-value}
  \label{algo:pval}
  \begin{algorithmic}[1]
    \REQUIRE Response $y$, design matrix $X$, index $g$ of next group to enter
    \ENSURE P-value for testing if $g$ and all other remaining groups are noise
    \STATE $\lambda \gets \norm{X_g^Ty}_2/w_g$, $r \gets \text{rank}(X_g)$
%    \IF{$X_g$ is a single column}
%    \STATE $P_g \gets 0$ is an $n \times n$ matrix
%    \STATE $\sigma^2_g \gets X_g^T \Sigma X_g$
%    \ELSE
    \STATE Compute $P_g$, $\sigma^2_g$ from equations \ref{eq:proj} and
      \ref{eq:cvar}
%    \ENDIF
    \RETURN $A$
  \end{algorithmic}
\end{algorithm}

@@

Since we are interested in variable selection we generate signals with
nonzero coefficients on the scale of $\gamma := \sqrt{2 \log(p)/n}$.
(\todo Cite some paper on support recovery?). Coefficients within a
nonzero group have roughly the same magnitude, and the magnitudes for
each nonzero group range from a lower limit times $\gamma$ to an upper
limit times $\gamma$. In plots, the limits are listed next to ``beta:''
and the numbers in parentheses are the largest and smallest 2-norms of
nonzero coefficient groups respectively. Each plot also displays the
number of observations or rows, $n$, the number of columns, $p$, the
number of groups, $g$, and the largest and smallest group sizes in
parentheses if the groups are not all size 1. The number of nonzero
groups, $k$, is also displayed via inward-facing tick marks on the
horizontal axes. The proportion of truly
nonzero groups recovered in the first $k$ steps is denoted $k$-SP,
where recall $k$ is the number of truly nonzero groups.

@@

\begin{figure}[h]
\begin{center}
\subfigure[Gaussian design]{
\label{fig:smallgauss}
\includegraphics[width=0.5\textwidth]{../figs/gaussian_n100_p200_g200_k5_lower1pt5_upper2_corr0pt2_noisecorr0pt1.pdf}}
\hspace{-15pt}
\subfigure[Categorical design]{
\label{fig:smallcat}
\includegraphics[width=0.5\textwidth]{../figs/categorical_n100_p100_g20_k4_lower4_upper5.pdf}}
\caption{\small \it The left panel shows results for Gaussian design
matrices with equicorrelation $\rho=0.2$ and equicorrelated noise with
correlation 0.1. The right panel shows results for categorical matrices
with all categories having 5 levels.}
\end{center}
\end{figure}

@@

\begin{figure}[h]
\begin{center}
\subfigure[Large Gaussian lasso]{
\label{fig:largelasso}
\includegraphics[width=0.5\textwidth]{../figs/gaussian_size5-10_n100_p200_g35_k10_lower1p8_upper1p9.pdf}}
\hspace{-15pt}
\subfigure[Large Gaussian group lasso]{
\label{fig:largegrplasso}
\includegraphics[width=0.5\textwidth]{../figs/gaussian_size5-10_n100_p200_g35_k10_lower1p8_upper1p9_corr0p1.pdf}}
\caption{\small \it The left panel shows results of a simulation with
  an independent Gaussian design matrix and a signal vector having
  three groups with non-zero coefficients of sizes one, two, and
  three, and seven groups of noise variables with sizes varying from
  one to ten, for a total of 30 variables. For the right panel we
  repeated this setup, but group sizes corresponded to levels of
  categorical variables. The groups of size one were increased to two,
so there are 10 categorical variables with a total of 32 levels.}
\end{center}
\end{figure}

@@

The large simulation has 50 groups, 25 of size 1, 10 of size 5, 10 of
size 10, and 5 of size 15. Signal vectors were supported on random
choices of 10 of these groups (changing in each realization), with
non-zero signal magnitude around $\sqrt{2\log p}$ where $p = 250$
(roughly 3.3). The design matrix had 200 rows and the simulation
performed 100 realizations.

@@

\appendix
\section{Derivation of closed forms of statistic}

>>>>

\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\renewcommand{\algorithmiccomment}[1]{\# #1}

@@

For various sparsity levels $k$, Algorithm~\ref{algo:fs} was run on
various data sets with \textit{steps} equal to $k$ and the ``average
power'' was calculated as the number of true variables in the active
set after $k$ steps divided by $k$. (\todo Make this consistent with
$\gamma$ in simulations section)
Nonzero coefficients had magnitude
in a range of multiples of $\gamma := \sqrt{2\log(G)}$, e.g. in $[1.4
\gamma, 1.6 \gamma]$. Results are shown in
Figure~\ref{fig:fwdstepsim}. After computing the
coherence of these design matrices, some calculations show the required
sparsity level to guarantee exact recovery in these situations is
about 2 or smaller, and the required nonzero coefficient magnitude is
likely in the range of 10 to 100 times $\gamma$. These simulations are far
from the stringent conditions required by theory to guarantee perfect
recovery, but the performance, while not perfect, may still be good
enough for some purposes.

@@

We now describe how to calculate our p-value, following the discussion of
group lasso in \cite{tests:adaptive}. For the rest of this section let
$g = \gstar$ be the index of
the group attaining the maximum on line 3 of Algorithm~\ref{algo:fs}.
%\lambda = \norm{X_g^Ty}_2/w_g$ be the value of the maximum, and
%$r = \text{rank}(X_g)$ be the rank of the maximizing group.
For a vector $u \in \real^p$, let $u_h$ denote the coordinates
of $u$ corresponding to the columns of group $h$ in the design matrix $X$.
We can rearrange the columns of $X$ to group these adjacently, so that
\[
u^TX^T = \begin{pmatrix} u_1^T X_1^T & u_2^T X_2^T & \cdots & u_G^TX_G^T \end{pmatrix}
\]
One step of the calculation will be to find an orthonormal basis for the
linear space $\vecsp_g = \{ u \in \real^p : u_g^T X_g^T y = 0, u_h = 0
\text{ for all } h \neq g \}$ so that we can project orthogonally to
this space. If $X_g$ is a single column and $X_g^Ty \neq 0$ (which
should be the case since $g$ maximizes the absolute value of this
quantity), then the
space is trivial and the desired orthogonal projection is the identity.
Otherwise, if $X_g$ has $p_g > 1$ columns the space $\vecsp_g$ will
generally have dimension $p_g-1$.
% have dimension $r-1$ for $r = \text{rank}(X_g)$. This holds except in
% degenerate cases such as $X_g = 0$ or $r \ll p_g$, but these cases are not
% likely to occur since $\norm{X_g^T y}_2/w_g$ is supposed to be maximized.
% We proceed by forming the $n \times p_g$ matrix
% $X_g - X_g X_g^T y y^T X_g/\norm{X_g^Ty}_2^2$ and performing Gram-Schmidt
% on the first $r-1$ columns. For precise theorems showing that this will
% generally work, see \cite{outerp:indep}.
We compute an orthonormal basis by Gram-Schmidt and form a
$p_g \times p_g$ matrix, which we denote $V_g$, by appending one column
%0's in the coordinates corresponding to all groups $h \neq g$ and to
of zeroes (since Gram-Schmidt only produces $p_g-1$ columns).
With this we are ready to define the projection

\begin{equation}
  \begin{aligned}
    \label{eq:proj}
    P_g &= \Sigma X_gV_g (V_g^T X_g^T \Sigma X_g V_g)^\dagger V_g^TX_g^T \\
   \end{aligned}
\end{equation}
Also define $H_g = (I-P_g)\Sigma$ and the conditional variance

\begin{equation}
  \begin{aligned}
    \label{eq:cvar}
    \sigma^2_g &= y^TX_gX_g^T H_g X_gX_g^Ty / (w_g^2 \norm{X_g^Ty}_2^2) \\
  \end{aligned}
\end{equation}
These simplify when $X_g$ is a single column, in which case $P_g = 0$,
$H_g = \Sigma$, and $\sigma^2_g = X_g^T\Sigma X_g^T/w_g^2$. We now
describe how to compute our p-value in Algorithm~\ref{algo:pval}. Note
that $F_{\chi^2_r}$ denotes the distribution function of a $\chi^2$
random variable with $r$ degrees of freedom.
%Define $U_g = X_g^Ty/(w_g\norm{X_g^Ty}_2)$.

\begin{algorithm}
  \caption{Computing p-value}
  \label{algo:pval}
  \begin{algorithmic}[1]
    \REQUIRE Response $y$, grouped design matrix $X$ with weights, inactive set $A^c$, index $g$ of last group to enter active set $A$.
    \ENSURE Our p-value for the group $g$ entering the model.
    \STATE Compute $H_g$ and $\sigma^2_g$
    \IF{$X_g$ is a single column}
    \STATE $\sigma_g^2 \gets X_g^T\Sigma X_g/w_g^2$
    \STATE $\tilde X_g \gets X_g / w_g \cdot \text{sign}(X_g^Ty)$
    \ELSE
    \STATE $P_g \gets \Sigma XV_g (V_g^T X^T \Sigma X V_g)^\dagger V_g^TX^T$
    \STATE $H_g \gets (I - P_g)\Sigma$
    \STATE $\sigma^2_g \gets y^TX_gX_g^T H_g X_gX_g^Ty / (w_g^2 \norm{X_g^Ty}_2^2)$
    \STATE $\tilde X_g \gets H_g X_g X_g^T y /(\norm{X_g^Ty}_2 w_g)$
    \ENDIF
    \STATE $\lambda \gets \norm{X_g^Ty}_2/w_g$, $r \gets \text{rank}(X_g)$
%    \STATE \COMMENT{Compute the following two $p$-vectors}
    \STATE $a \gets X^T(y - \lambda \tilde X_g)$
    \STATE $b \gets X^T \tilde X_g$
    \STATE Compute solution $(v_-, v_+)$ of \textsc{LinFrac}(a,b) optimization subproblem
    \STATE $u \gets [F_{\chi^2_r}(v^2_-/\sigma_g^2) - F_{\chi^2_r}(\lambda^2/\sigma_g^2)]/[F_{\chi^2_r}(v^2_-/\sigma_g^2) - F_{\chi^2_r}(v^2_+/\sigma_g^2)]$
    \RETURN $u$
  \end{algorithmic}
\end{algorithm}

We next describe the \textsc{LinFrac} subproblem and its solution. We
gave the problem this name because it originated in the form
\[
\maximize_{h \neq g, \norm{u_h}_2 = 1} \frac{u_h^TX_h^Ty - u_h^TX_h^TX_gX_g^Ty}
{1 - u_g^TX_g^TX_hu_h}
\]
The solution we describe next is to a slightly different problem, and
also incorporates the information that $g$ maximizes $\norm{X_g^Ty}_2/w_g$.
For details see Appendix~\ref{app:closedform} or \cite{tests:adaptive}.

\begin{algorithm}
  \caption{The \textsc{LinFrac} optimization subproblem}
  \label{algo:linfrac}
  \begin{algorithmic}[1]
    \REQUIRE The $p$-vectors $a, b$ in Algorithm~\ref{algo:pval}, weights, inactive set $A^c$, a small tolerance number
%(we take $1e^{-10}$)
    \ENSURE Solution pair $(v_-, v_+)$.
    \FOR{$h$ in $A^c$}
      \IF{$\norm{b_h}_2 == 0$ or $\norm{a_h}_2/\norm{b_h}_2 < \text{tol}$}
        \STATE $(v^-_h, v^+_h) \gets (0, \infty)$
      \ELSE
        \STATE $\theta_c \gets a_h^Tb_h/(\norm{a_h}_2 \norm{b_h}_2)$
        \STATE $\theta_s \gets \sqrt{1-\theta_c^2}$
        \STATE $\theta \gets \arccos (\theta_c)$
        \STATE $\phi_s \gets \theta_s \norm{b_h}_2 / w_h$
        \IF{$\phi_s > 1$}
          \STATE $(v^-_h, v^+_h) \gets (0, \infty)$
        \ELSE
          \STATE $\phi \gets \arcsin(\phi_s)$
%          \STATE $\phi_2 \gets \pi - \phi_1$
          \STATE $z_\pm \gets s_\pm \norm{a_h}_2 \cos(\phi) /(w_h - s_\pm \norm{b_h}_2 \cos(\theta - \phi))$ for $s_\pm = \pm 1$
          \IF{$\norm{b_h}_2 < w_h$}
            \STATE $(v_h^-, v_h^+) \gets (\max \{ z_+, z_- \}, \infty)$
          \ELSE
            \STATE $(v_h^-, v_h^+) \gets (\min \{ z_+, z_- \}, \max \{ z_+, z_- \})$
          \ENDIF
        \ENDIF
      \ENDIF
    \ENDFOR
    \STATE $v^- \gets \min_h v_h^-$
    \STATE $v^+ \gets \max_h v_h^+$
    \RETURN $(v^-, v^+)$
  \end{algorithmic}
\end{algorithm}

@@

Since we are interested in variable selection we generate signals with
nonzero coefficients on the scale of $\gamma := \sqrt{2 \log(p)/n}$.
(\todo Check this scaling. \todo Cite some paper on support recovery?).
Coefficients within a
nonzero group have roughly the same magnitude, and the magnitudes for
each nonzero group range from a lower limit times $\gamma$ to an upper
limit times $\gamma$. In plots, the limits are listed next to ``beta:''
and the numbers in parentheses are the largest and smallest 2-norms of
nonzero coefficient groups respectively. Each plot also displays the
number of observations or rows, $n$, the number of columns, $p$, the
number of groups, $g$, and the largest and smallest group sizes in
parentheses if the groups are not all size 1. The number of nonzero
groups, $k$, is also displayed via inward-facing tick marks on the
horizontal axes. The proportion of truly
nonzero groups recovered in the first $k$ steps is denoted $k$-SP,
where recall $k$ is the number of truly nonzero groups.

@@

\begin{figure}[h]
\begin{center}
\subfigure[Gaussian with correlations]{
\label{fig:gausscorr}
\includegraphics[width=0.5\textwidth]{../figs/gaussian_n100_p200_g200_k5_lower1pt5_upper2_corr0pt2_noisecorr0pt1.pdf}}
\hspace{-15pt}
\subfigure[Gaussian with groups]{
\label{fig:gaussgroups}
\includegraphics[width=0.5\textwidth]{../figs/gaussian_n100_p200_g35_k5_lower2_upper2pt5.pdf}}
\caption{\small \it The left panel shows results for Gaussian design
matrices with equicorrelation $\rho=0.2$ and equicorrelated noise with
correlation 0.1. The right panel shows results for an independent
Gaussian design with groups of sizes 5 and 10.}
\end{center}
\end{figure}

@@

\begin{figure}[h]
\begin{center}
\subfigure[Real data categorical design]{
\label{fig:hivnrti}
\includegraphics[width=0.5\textwidth]{../figs/HIV_NRTI_n848_p48_g14_k4_lower21_upper25.pdf}}
\hspace{-15pt}
\subfigure[Simulated categorical design]{
\label{fig:gausscat}
\includegraphics[width=0.5\textwidth]{../figs/categorical_n100_p100_g20_k4_lower4_upper5.pdf}}
\caption{\small \it The left panel shows results for a categorical design
taken from a real data set.
The right panel shows results for categorical matrices with all categories having 5 levels.}
\end{center}
\end{figure}

@@

@@

\appendix
\section{Derivation of closed forms of statistic}
\label{app:closedform}
